Proof-Reconstruction in Type Theory for Propositional Logic

### 2017-10-23 paper/sections/proof-reconstruction.tex
[ paper ] fixing lemmas.
[ paper ] working on resolve.
[ paper ] diagram fixed.

@@

The proof-reconstruction approach proposed consists of a series
of steps similar to the workflow presented by \citeauthor{sultana2015} in
\cite{sultana2015}. This process is a translation from a source
system to a target system. In our case, the system of origin, the
automatic theorem prover, is \Metis; the target system is a proof
assistant, \Agda. We choose \Agda, but another proof assistant with the
same support of type theory and inductive types could be used.
%  (see a description of \Agda for its main features in
% Section~\ref{sec:implementation}).
The overview of the proof-reconstruction is presented in
Fig.~\ref{fig:proof-reconstruction-workflow}. The process begin with
a \TPTP file that encodes a problem in \CPL. We use this file as the
input of the \Metis prover and if the problem is a theorem, \Metis can
reply a derivation of the proof in \TSTP format.

@@

Finally, we type-check the \Agda proof-term. If the type-checking
success, the \TSTP derivation delivered by \Metis is correct
module \Agda and the proposed formalizations for the
propositional logic and the \Metis inferences.
In that case \Agda outputs an interface file.
Otherwise, when type-checking fails, the failure must be
investigated by the user looking at the error in the \TSTP derivation
by \Metis, in the translation by \Athena, in the \Agda formalizations
mentioned above or even in the type-checker~\Agda.

In the remainder part of this section, a formal description using
type theory is presented to build definitions and theorems of
functions necessary to emulate \Metis' inference rules.
As a convention, we write the type judgments as ``$m : M$'' to denote
that $m$ has $\abbre{M}$ type.
Types are written using small caps.
We use \Prop type for propositions and the dependent type, $Γ ⊢ φ$ for
theorems where $Γ$ has $\abbre{List}~\Prop$ type, and $φ$ has type $\Prop$.
$\abbre{List}$ type is the usual inductive type for lists.
For syntactical equivalence, we write the equality as $φ ≡ ψ$
for $φ : \Prop$ and $ψ : \Prop$. For this equality, we assume the
reflexivity, symmetry and transitivity property.

% ...................................................................

@@

To emulate \Metis' inference rules in type theory, we observed that
transformations performed by some rules are given by \emph{general recursive}
functions. Algorithms based on general recursions can not be translated
directly in type theory since it is not a guarantee they terminate. For that
reason, we follow the technique described in~\cite{Bertot2004} to avoid
termination problems by modifying the recursive functions to be
\emph{structural recursive}.
% (see \cite{Coquand1992,Abel2002,Bove2005} for more details).

@@

General recursive calls can be translated into structurally recursive
calls by using for instance, the \emph{Bounded Recursion} technique
presented in~\cite{Bertot2004}. For a yet another method, we refer
the reader to \cite{Coquand1992,Abel2002,Bove2005}.
The former technique defines a new function based on the original
recursive function by adding an argument to the function.
The new argument is the \emph{bound}, a natural number given by the
function complexity. In other words, the added argument will represent
the number of times the function needs to call itself to get the
expected outcome~\cite{Bertot2004}.

The type of natural numbers is called \Nat type, and it is defined as
usual, \ie \zero and \suc are the data constructors. We use names
and symbols of the arithmetic operations as usual.
We include syntax sugar for $\zero$, $\suc$, $\suc(\zero)$,
$\cdots$, with the decimal representation: $0$, $1$, $2$, $\cdots$
as well.

Therefore, for a general recursive function $\rm{f}: \Prop \to \Prop$
with termination problems, we formulate a new function
$\rm{f}^{*} : \Prop \to \Nat \to \Prop$ where all recursive calls must be
done by using induction on the \Nat type structure.

@@

Let us consider the following example to exhibit the bounded technique
mentioned above for the \fuh function presented in the proposed version
for the \Metis \strip inference.

\begin{equation*}
\label{eq:uh-definition}
\begin{aligned}
&\hspace{.495mm}\fuh_{0} : \Prop → \Prop\\
&\begin{array}{lll}
  \fuh_{0} &(φ₁ ⇒ (φ₂ ⇒ φ₃)) &= \fuh_{0}((φ₁ ∧ φ₂) ⇒ φ₃)\\
  \fuh_{0} &(φ₁ ⇒ (φ₂ ∧ φ₃)) &= \fuh_{0}(φ₁ ⇒ φ₂) ∧ \fuh_{0}(φ₁ ⇒ φ₃)\\
  \fuh_{0} &φ                &= φ
\end{array}
\end{aligned}
\end{equation*}

From the definition above, the size of the argument, $(φ₁ ∧ φ₂) ⇒ φ₃$,
is not structurally smaller than the input formula, $φ₁ ⇒ (φ₂ ⇒ φ₃)$,
since $(φ₁ ∧ φ₂) ⇒ φ₃$ is not a subformula of $φ₁ ⇒ (φ₂ ⇒ φ₃)$.

Therefore, we define the function
$\fuh_{1}$ by structural recursion in its second argument in
Eq.~(\ref{eq:uh-structured}).

\begin{equation}
\label{eq:uh-structured}
\begin{aligned}
&\hspace{.495mm} \fuh_{1} : \Prop → \Nat \to \Prop\\
&\begin{array}{llll}
\fuh_{1} &(φ₁ ⇒ (φ₂ ⇒ φ₃)) &(\suc~n) &= \fuh_{1}~((φ₁ ∧ φ₂) ⇒ φ₃)~n\\
\fuh_{1} &(φ₁ ⇒ (φ₂ ∧ φ₃)) &(\suc~n) &= \fuh_{1}~(φ₁ ⇒ φ₂)~n ∧ \fuh_{1}~(φ₁ ⇒ φ₃)~n\\
\fuh_{1} &φ &\xspace\text{n} &= φ
\end{array}
\end{aligned}
\end{equation}

@@

\begin{equation*}
% \label{eq:rank-definition}
\begin{aligned}
&\hspace{.495mm}\fuh_{cm} : \Prop → \Nat\\
&\begin{array}{llll}
\fuh_{cm}~(φ₁ ⇒ (φ₂ ⇒ φ₃)) &= \fuh_{cm}(φ₃) + 2\\
\fuh_{cm}~(φ₁ ⇒ (φ₂ ∧ φ₃)) &= \fmax~(\fuh_{cm}~φ₂)~(\fuh_{cm}~φ₃) + 1\\
\fuh_{cm}~φ                &= 0
\end{array}
\end{aligned}
\end{equation*}

Finally, following the technique mentioned, we get a final version of
$\fuh_{0}$ function presented in Eq.~(\ref{eq:uh}).

\begin{align}
  \label{eq:uh}
  \begin{split}
  &\fuh : \Prop → \Prop\\
  &\fuh(φ) = \fuh_{1}(φ, \fuh_{cm}(φ)).
  \end{split}
\end{align}

@@

\subsection{Emulating \Metis' inference rules}
\label{ssec:emulating-inferences}

In this section, we emulate each \Metis inference rule with a
function and its respective theorem.
The following pattern is used to present the formalization.
The function name written in bold font
is a \Metis inference rule in \TSTP derivations.
The same function name written in roman letters refers to the proposed
version of the rule.
Each rule proposed has a respective theorem also written in roman letters
that proves its soundness.

@@

\begin{verbatim}
fof(normalize_1_2, plain, (p | r),
    inference(conjunct, [], [normalize_1_1])).
\end{verbatim}

For this inference rule, we proposed the \rm{conjunct} function in
Def.~\ref{def:conjunct} and we proved Thm.~\ref{thm:thm-conjunct}. The pattern is the following for
a binary rule and we extend it for $n$-ary rules in a natural way.

\begin{itemize}
\item \texttt{rule} is a binary \Metis inference rule.
\item $\rm{rule}$ is the proposed function for the \texttt{rule} inference.
\begin{align*}
  \begin{split}
    &\rm{rule} : \Prop → \Prop → \Prop\\
    &\rm{rule}\ φ₁\ φ₂\ ψ\ =
      \begin{cases}
      ψ, &\text{if }\rm{rule}\text{ built }ψ\text{ by applying valid transformations}\\
         &\text{or other rules to }φ₁\text{ and }φ₂\\
      φ, &\text{otherwise.}
      \end{cases}
  \end{split}
\end{align*}

\begin{itemize}
  \item $φᵢ$ are \emph{source} formulas from previous deducted derivations.
  \item $ψ$ is the \emph{target} formula. The expected result of \texttt{rule}.
\end{itemize}
\item The theorem that proves the soundness of the
$\rm{rule}$ function looks like
\begin{equation*}\text{If }Γ ⊢ φ\text{ then }Γ ⊢ \rm{rule}\ φ.\end{equation*}

\end{itemize}

\begin{table}[!ht]
  \begin{center}
  {\renewcommand{\arraystretch}{1.6}%
    \begin{tabular}{|@{\hspace{2mm}}l@{\hspace{2mm}}l@{\hspace{2mm}}l@{\hspace{2mm}}|}
    \hline
    \textbf{\Metis rule} & \textbf{Purpose} &\textbf{Theorem}\\ \hline
    \texttt{canonicalize}
      &Normalization of the formula
      &Thm.~\ref{thm:canonicalize}
    \\
    \texttt{clausify}
      &Performs clausification
      &Thm.~\ref{thm:clausify}
    \\
    \hyperlink{atp-conjunct}{\texttt{conjunct}}
      &Takes a formula from a conjunction
      &Thm.~\ref{thm:conjunct}
    \\
    % \texttt{negate}
    %   &Negate the formula
    %   &Thm.~\ref{thm:negate}
    % \\
    \texttt{resolve}
      &A general form of the resolution theorem
      &Thm.~\ref{thm:resolve}
    \\
    \texttt{simplify}
      &Simplify definitions and theorems
      &Thm.~\ref{thm:simplify}
    \\
    \texttt{strip}
      &Strip a goal into subgoals
      &Thm.~\ref{thm:strip}
    \\[1ex]
    \hline
    \end{tabular}}
  \end{center}
\label{tab:agda-metis-table}
\end{table}

@@

\subfile{sections/proof-reconstruction/strip.tex}
\subfile{sections/proof-reconstruction/negate.tex}
\subfile{sections/proof-reconstruction/conjunct.tex}
% \subfile{sections/proof-reconstruction/resolve.tex}
% \subfile{sections/proof-reconstruction/canonicalize.tex}
\subfile{sections/proof-reconstruction/clausify.tex}
\subfile{sections/proof-reconstruction/simplify.tex}

>>>>

MISSING TEXT

@@

The proof-reconstruction approach proposed consists of a series
of steps similar to the workflow presented by \citeauthor{sultana2015} in
\cite{sultana2015}. This process is a translation from a source
system to a target system. In our case, the system of origin, the
automatic theorem prover, is \Metis; the target system is a proof
assistant, \Agda. We choose \Agda, but another proof assistant with the
same support of type theory and inductive types could be used.
%  (see a description of \Agda for its main features in
% Section~\ref{sec:implementation}).
The overview of the proof-reconstruction is presented in
Fig.~\ref{fig:proof-reconstruction-workflow}. The process begin with
a \TPTP file that encodes a problem in \CPL. We use this file as the
input of the \Metis prover. If the problem is a theorem, \Metis
reply a derivation of the proof in \TSTP format.

@@

Finally, we type-check the \Agda proof-term. If the type-checking
success, the \TSTP derivation delivered by \Metis is correct
module \Agda and the proposed formalizations for the
propositional logic and the \Metis inferences.
In that case \Agda outputs an interface file.
Otherwise, when type-checking fails, the failure must be
investigated by the user looking at the error in the \TSTP derivation
by \Metis, in the translation by \Athena, in the \Agda formalizations
mentioned above or in the type-checker~\Agda.

In the remainder part of this section, a formal description using
type theory is presented to build definitions and theorems of
functions necessary to emulate \Metis' inference rules.

\begin{notation}
We write the type judgments as $m : \abbre{M}$ to denote
that $m$ has $\abbre{M}$ type.
Types are written using small caps.
We use \Prop for the type of propositions and the dependent type, $Γ ⊢ φ$ for
theorems where $Γ : \List~\Prop$ and $φ : \Prop$.
$\abbre{List}$ type is the usual inductive type for lists.
For syntactical equivalence, we write the equality as $φ ≡ ψ$
for $φ,~ψ : \Prop$. For this equality, we assume the
reflexivity, symmetry and transitivity properties.
\end{notation}
% ...................................................................

@@

To reconstruct \Metis' inference rules in type theory, we observed that
transformations performed by some rules are implemented by \emph{general
recursive} functions.
Algorithms based on general recursions can not be translated
directly in type theory since it is not a guarantee they terminate. For that
reason, we follow the technique described in~\cite{Bertot2004} to avoid
termination problems by modifying the recursive functions to be
\emph{structural recursive}.
% (see \cite{Coquand1992,Abel2002,Bove2005} for more details).

@@

General recursive functions can be translated into structurally recursive
functions by using for instance, the \emph{Bounded Recursion} technique
presented in~\cite{Bertot2004}. For a yet another methods, we refer
the reader to \cite{Coquand1992,Abel2002,Bove2005}.
The bounded technique defines a new structural recursive function based on
the general recursive function by adding an argument to the function.
The new argument is the \emph{bound}, a natural number given by the
function complexity. In other words, the added argument will represent
the number of times the function needs to call itself to get the
expected outcome~\cite{Bertot2004}.

\begin{notation}
The type of natural numbers is called \Nat, and it is defined as
usual, where \zero and \suc are its data constructors. We use names
and symbols of the arithmetic operations as usual.
We include syntax sugar for $\zero$, $\suc$, $\suc(\zero)$,
$\cdots$, with the decimal representation: $0$, $1$, $2$, $\cdots$
as well.
\end{notation}

Therefore, one approach to guarantee a structural recursive function based on
a general recursive function $\rm{f}: \Prop \to \Prop$ is to formulate a new function $\rm{f}^{*} : \Prop \to \Nat \to \Prop$ where all recursive calls are done on smaller arguments.

@@

Let us consider the following example to exhibit the bounded technique
mentioned above for the \fuh function. this function is used for reconstructing
the \Metis \strip inference rule in Section~\ref{sssec:strip-a-goal}.

\begin{equation}
\label{eq:uh-definition}
\begin{aligned}
&\hspace{.495mm}\fuh_{0} : \Prop → \Prop\\
&\begin{array}{lll}
  \fuh_{0} &(φ₁ ⇒ (φ₂ ⇒ φ₃)) &= \fuh_{0}~((φ₁ ∧ φ₂) ⇒ φ₃)\\
  \fuh_{0} &(φ₁ ⇒ (φ₂ ∧ φ₃)) &= \fuh_{0}~(φ₁ ⇒ φ₂) ∧ \fuh_{0}~(φ₁ ⇒ φ₃)\\
  \fuh_{0} &φ                &= φ
\end{array}
\end{aligned}
\end{equation}

In (\ref{eq:uh-definition}), the first two equation show
call on arguments no structurally smaller.
For instance, on the first call, $φ ≡ (φ₁ ∧ φ₂) ⇒ φ₃$, the recursion
calls on the  $φ₁ ⇒ (φ₂ ⇒ φ₃)$ argument.
But this argument is not a subformula for φ.
Therefore, one way to translate $\fuh_{0}$ in (\ref{eq:uh-definition}) to type
theory is to define the a new function using the bounded technique.
\begin{equation}
\label{eq:uh-structured}
\begin{aligned}
&\hspace{.495mm} \fuh_{1} : \Prop → \Nat \to \Prop\\
&\begin{array}{llll}
\fuh_{1} &(φ₁ ⇒ (φ₂ ⇒ φ₃)) &(\suc~n) &= \fuh_{1}~((φ₁ ∧ φ₂) ⇒ φ₃)~n\\
\fuh_{1} &(φ₁ ⇒ (φ₂ ∧ φ₃)) &(\suc~n) &= \fuh_{1}~(φ₁ ⇒ φ₂)~n ∧ \fuh_{1}~(φ₁ ⇒ φ₃)~n\\
\fuh_{1} &φ &\xspace\text{n} &= φ
\end{array}
\end{aligned}
\end{equation}

@@

\begin{equation*}
  \begin{aligned}
    &\hspace{.495mm}\fuh_{cm} : \Prop → \Nat\\
    &\begin{array}{llll}
    \fuh_{cm}~(φ₁ ⇒ (φ₂ ⇒ φ₃)) &= \fuh_{cm}~φ₃ + 2\\
    \fuh_{cm}~(φ₁ ⇒ (φ₂ ∧ φ₃)) &= \fmax~(\fuh_{cm}~φ₂)~(\fuh_{cm}~φ₃) + 1\\
    \fuh_{cm}~φ                &= 0
    \end{array}
  \end{aligned}
\end{equation*}

Following the technique mentioned above, we get the structural recursive definition of $\fuh_{0}$ function presented in ~(\ref{eq:uh}).

\begin{align}
  \label{eq:uh}
  \begin{split}
  &\fuh : \Prop → \Prop\\
  &\fuh~φ~ = \fuh_{1}~φ~(\fuh_{cm}~φ).
  \end{split}
\end{align}

@@

\subsection{Reconstructing \Metis' inference rules}
\label{ssec:emulating-inferences}

In this section, we reconstruct each \Metis inference rule.
We present a function and a theorem for each rule.

\begin{notation}
The function name written in bold font is a \Metis inference rule.
The same function name written in roman letters refers to the proposed
version of the rule.
Each rule proposed has a respective theorem also written in roman letters
that proves its soundness.
\end{notation}

The following pattern is used to present the reconstruction.

@@

\begin{verbatim}
fof(norm1, (p | r), inf(conjunct, [norma1])).
\end{verbatim}

For this inference rule, we the \rm{conjunct} function in~(\ref{def:conjunct}) and we proved Theorem~\ref{thm:thm-conjunct}. The pattern is the following for
a binary rule and we extend it for $n$-ary rules in a natural way.

\begin{itemize}
\item \conjunct is a binary \Metis inference rule.
\item $\rm{conjunct}$ is the type theory function for the \texttt{rule} inference.
\begin{align*}
  \begin{split}
    &\rm{conjunct} : \Prop → \Prop → \Prop\\
    &\rm{conjunct}\ φ₁\ φ₂\ ψ\ =
      \begin{cases}
      ψ, &\text{if }\rm{conjunct}\text{ built }ψ\text{ by applying valid transformations}\\
         &\text{or other rules to }φ₁\text{ and }φ₂\\
      φ, &\text{otherwise.}
      \end{cases}
  \end{split}
\end{align*}

\begin{itemize}
  \item $φᵢ$ are \emph{source} formulas from previous deducted derivations.
  \item $ψ$ is the \emph{target} formula. The expected result of \texttt{rule}.
\end{itemize}
\item The theorem that proves the soundness of the
$\rm{conjunct}$ function would be
\begin{equation*}\text{If }Γ ⊢ φ\text{ then }Γ ⊢ \rm{conjunct}\ φ.\end{equation*}
\end{itemize}

\begin{table}[!ht]
\label{tab:agda-metis-table}
\caption{Presentation of \Metis inference rules in type theory.}
  \begin{center}
  {\renewcommand{\arraystretch}{1.6}%
    \begin{tabular}{|@{\hspace{2mm}}l@{\hspace{2mm}}l@{\hspace{2mm}}l@{\hspace{2mm}}|}
    \hline
    \textbf{\Metis rule} & \textbf{Purpose} &\textbf{Theorem}\\ \hline
    \texttt{canonicalize}
      &Normalization of the formula
      &Theorem~\ref{thm:canonicalize}
    \\
    \texttt{clausify}
      &Performs clausification
      &Theorem~\ref{thm:clausify}
    \\
    \hyperlink{atp-conjunct}{\texttt{conjunct}}
      &Takes a formula from a conjunction
      &Theorem~\ref{thm:conjunct}
    \\
    % \texttt{negate}
    %   &Negate the formula
    %   &Theorem~\ref{thm:negate}
    % \\
    \texttt{resolve}
      &A general form of the resolution theorem
      &Theorem~\ref{thm:resolve}
    \\
    \texttt{simplify}
      &Simplify definitions and theorems
      &Theorem~\ref{thm:simplify}
    \\
    \texttt{strip}
      &Strip a goal into subgoals
      &Theorem~\ref{thm:strip}
    \\[1ex]
    \hline
    \end{tabular}}
  \end{center}
\end{table}

MISSING TEXT

@@

\subfile{sections/proof-reconstruction/strip.tex}
\subfile{sections/proof-reconstruction/negate.tex}
\subfile{sections/proof-reconstruction/conjunct.tex}
\subfile{sections/proof-reconstruction/resolve.tex}
\subfile{sections/proof-reconstruction/canonicalize.tex}
\subfile{sections/proof-reconstruction/clausify.tex}
\subfile{sections/proof-reconstruction/simplify.tex}
